NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-06-19 Manuscript/manuscriptPS.tex

Previous experimental studies have concluded that
PS headgroups have more rigid conformation than
phophocholines (PC) due to hydrogen bonding or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations or Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with PS
containing bilayers \cite{hauser77,kurland79,hauser83,hauser85,feigenson86,mattai89,roux90,roux91}.
Dilution of PS rich bilayers with PC lipids
reduces the rigidity and propensity of the
strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.

@@

    DOPS/Na$^+$  & CHARMM36 \cite{??}  \todoi{Correct citation for CHARMM DOPS}      &0 & 0 & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\

@@

    POPS/Na$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}&0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
    POPS/K$^+$   & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}&0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\

@@

\caption{  List of MD simulations. The salt concentrations calculated as 
   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{??})
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,??} \todoi{Correct citation for CHARMM POPS}  &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{??}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{??}          &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
     POPC:POPS (5:1)        & CHARMM36 \cite{??}\todoi{These are with NBFIX from Ref. \citenum{kim16}}          &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS (5:1)        & CHARMM36 \cite{??}\todoi{These are with NBFIX from Ref. \citenum{kim16}}          &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS\todoi{This is also probably OPPS? These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
   POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &300 & 120:24 & 5760 & 31  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &1000 & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    \todo{MacRog simulations with KCl to be added}\\
    \todo{Berger simulations with NaCl and CaCl to be added}

@@

In order to compare the sodium interaction with PS containing membranes between
experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the headgroup
order parameters of POPC and POPS measured from 5:1 mixture as a
function of different monovalent ions added to the buffer.
The experimental experimental results \cite{roux86} are compared with different simulation
models.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{Maybe we should get the CHARMM results as well?}
\end{figure*}
In line with the trend observed for bilayers in the absence of negatively
charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
significantly larger than to potassium, indicating stronger binding of lithium
ions to lipid bilayers. The unique nature of lithium is also evident in the
response of PS headgroup order parameters to bounds ions, which decrease
with the addition of lithium or divalent cations but increase with the
addition of sodium or potassium. The headgroup order parameter response
to the added NaCl in the Berger model is similar to the experiments of LiCl,
except for the $\alpha$ carbon of POPS for which the change is overestimated
with the order of magnitude. The results suggests that the binding affinity
of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
and that the response of $\alpha$ carbon of POPS to the added sodium
is significantly overestimated. The response of POPC to the added potassium
in MacRog model is only slightly overestimated, in contrast to the response
of POPS where order of magnitude overestimations of reposonse are observed. 
In conclusion, it seems that the response of PS headgroup to bound cations
is not correctly captured in the tested models.
\todo{Finish the discussion when more models are available}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
  \todo{Fix CKP1 and CKP2 captions.}
\end{figure}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
\end{figure*}

>>>>

Previous experimental studies have concluded that
PS headgroups have more rigid conformation than
phophocholines (PC) due to hydrogen bonding or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with PS
containing bilayers \cite{hauser77,kurland79,hauser83,hauser85,feigenson86,mattai89,roux90,roux91}.
Dilution of bilayers with PC lipids makes PS headgroups
less rigid and reduces propensity for the formation of
strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.

@@

    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       &0 & 0 & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\

@@

    POPS/Na$^+$  & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
    POPS/K$^+$   & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\

@@

\begin{table*}[!htb]
%\begin{sidewaystable*}[!p]
\centering
\caption{  List of MD simulations. The salt concentrations calculated as 
   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{??})
   and CKP refers to the version with more Gromos compatible version
   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
   %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
   %   The lipid force fields named as in our previous work~\cite{botan15}.
}\label{IONsystems}
%begin{minipage}[t]{\textwidth}
\begin{tabular}{l c c r r r r r r c c}
  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
  \hline
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}   &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}    &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    \hline
    POPC:POPS\todoi{This is also probably OPPS? These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
   POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &300 & 120:24 & 5760 & 31  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &1000 & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    \todo{MacRog simulations with KCl to be added}\\
    \hline
    \todo{Berger simulations with NaCl and CaCl to be added}
\end{tabular}
%\end{minipage}
%\end{sidewaystable*}

@@

In order to compare the sodium interaction with PS containing membranes between
experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the headgroup
order parameters of POPC and POPS measured from 5:1 mixture as a
function of different monovalent ions added to the buffer.
The experimental experimental results \cite{roux86} are compared with different simulation
models.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{CHARMM36 results for this plot would be highly useful.}
\end{figure*}
In line with the trend observed for bilayers in the absence of negatively
charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
significantly larger than to potassium, indicating stronger binding of lithium
ions to lipid bilayers. The unique nature of lithium is also evident in the
response of PS headgroup order parameters to bounds ions, which decrease
with the addition of lithium or divalent cations but increase with the
addition of sodium or potassium. The headgroup order parameter response
to the added NaCl in the Berger model is similar to the experiments of LiCl,
except for the $\alpha$ carbon of POPS for which the change is overestimated
with the order of magnitude. The results suggests that the binding affinity
of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
and that the response of $\alpha$ carbon of POPS to the added sodium
is significantly overestimated. The response of POPC to the added potassium
in MacRog model is only slightly overestimated, in contrast to the response
of POPS where order of magnitude overestimations of reposonse are observed. 
In conclusion, it seems that the response of PS headgroup to bound cations
is not correctly captured in the tested models.
\todo{Finish the discussion when more models are available}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
  \todo{Lipid 17 is to be added.}
\end{figure}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.}
\end{figure*}
